Mathematical Formulation of Projection Problem

### 2024-09-22 docs/projection.tex
Corrected objective function

\section{Goal}
To project a set of points $\mathcal X$ onto the nearest ellipsoid with volume 1, such that the projection of this ellipsoid in two different views produces ellipses which minimise the error between their areas and two target areas.

@@

\subsection{Sampling Approach}
A point-sampling-based approach is more involved, though is more relevant to the water balloon example which this toy example will graduate to.

After obtaining the projected ellipsoid with unit volume,

>>>>

\section{Goal}
To project a set of points $\mathcal X$ onto the nearest ellipsoid with volume 1, such that the projection of this ellipsoid in two different views produces ellipses which minimise the error to two target ellipses. This error can be measured as the error between the areas of the ellipses or their point clouds.

@@

\subsection{Sampling Approach}
A point-sampling-based approach is more involved, though is more relevant to the water balloon example which this toy example will graduate to. In this approach, we are no longer minimising the error between the ellipses' areas and their targets, but matching the ellipses themselves. As a result, we need to sample the points of the projected \textit{ellipses} and minimise the difference between this set of points and a target set of points (such as a set of points sampled from a circle).

In literature, there have been various proposed metrics to measure the difference between two sets of points. In particular, these metrics are important for point cloud registration, where the goal is to align multiple point clouds. Hausdorff distance has been used previously to evaluate the difference between two point clouds, however, as it is not traditionally differentiable, we will be using Chamfer distance. The Chamfer distance $D_C$ between two sets of points $\mathcal{X}$ and $\mathcal{Y}$ is defined as 
\begin{equation}
    D_C(\mathcal{X,Y})= \frac{1}{|\mathcal{X}|}\sum_{x \in \mathcal{X}} d(x, \mathcal Y) + \frac{1}{|\mathcal{Y}|}\sum_{y \in \mathcal{Y}} d(y, \mathcal X),
\end{equation}
where $d(a, \mathcal S)$ for a point $a$ and a set $\mathcal S$ is defined as $\min d(a,s)$ for all $s \in \mathcal S$. In other words, this gives the sum of mean distance between all pairs of points from $\mathcal X$ to $\mathcal Y$ and the opposite. As Euclidean distance is known to be sensitive to outliers, other variations like density-aware and hyperbolic Chamfer distance have been researched.

Conceptually, after projected the rotated unit volume ellipsoid onto the $x$-$y$ plane, we sample around these 2D ellipses. We then calculate the Chamfer distance of these sampled points to our ground truth points at each step and attempt to minimise this distance through backpropagation.

The matter of sampling the points poses issues if not done carefully as stochastic nodes cannot be backpropagated through. We take inspiration from variational autoencoders (VAE) and apply the "reparametrisation trick" such that the random variable can be expressed in terms of deterministic nodes through which the gradient can flow through and a stochastic node to perform the sampling.

Let the projected ellipses $\varepsilon_1$ and $\varepsilon_2$ be represented by the matrices $\hat{\mathbf A}_1$ and $\hat{\mathbf A}_2$, respectively, as above. For these ellipses to be well-defined, both these matrices must be symmetric positive definite. As proven by Horn \& Johnson, there exists a unique positive definite square root $B=A^{1/2}$ for a positive definite and symmetric matrix $A$ such that $A= A^{1/2} A^{1/2}  = BB$. Furthermore, as it is positive definite, $A^{1/2}$ is invertible -- that is, $A^{-1/2}$ exists.

Thus, the reparametrisation trick applied is to decompose $\hat{\mathbf A}_i$ into its square roots like so:
\begin{align}
    x^\top \hat{\mathbf A}_i x = 1 &\implies x^\top \hat{\mathbf A}^{1/2}_i \hat{\mathbf A}^{1/2}_i x = 1 \\
    &\implies z^\top z = 1,
\end{align}
where we let $z = \hat{\mathbf A}^{1/2}_i x$. Thus, we simply need to sample unit vectors $z$ (or vectors on the unit circle) and can recover the sampled point $x$ by applying the inverse square root. That is, $x = \hat{\mathbf A}^{-1/2}z$.

Let the sampled points of the projected ellipses be denoted by $\mathcal X_1(\mathbf u^*)$ and $\mathcal X_2(\mathbf u^*)$ for $\varepsilon_1$ and $\varepsilon_2$, respectively, using the reparametrisation trick. Hence, given two sets of ground truth points $\mathcal Y_1, \mathcal Y_2$ for the projected ellipses $\varepsilon_1$ and $\varepsilon_2$, our final minimisation problem is
$$
\begin{array}{ll}
    \underset{\text{over} \mathcal X = \{ \mathbf x_i \}}{\text{minimize}} &  d_C(\mathcal X_1(\mathbf u^*), \mathcal Y_1) + d_C(\mathcal X_2(\mathbf u^*), \mathcal Y_2) \\[0.5em]
  \text{subject to} & \mathbf u^* = \text{argmin} \; \lr\Vert\Vert {\text{diag}(\mathbf X^\top Q(\mathbf u_{4:6}) ^\top \Lambda(\mathbf u_{1:3}) Q(\mathbf u_{4:6}) \mathbf X) - \mathbf 1_m} ^2 \\[0.5em]
  & \hphantom{y =} \text{subject to} \;  \frac{4}{3} \pi \mathbf u_1 \mathbf u_2 \mathbf u_3 = 1
\end{array}.
$$
